Climate indicators

### 2023-02-22 DRAFT_climate_indicators.Rmd
trying to normalise

```{r}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
t1 = as.Date("2018-05-31")
x = read_stars(c(tif, tif, tif, tif), along = 
                  list(time = c(t1, t1+1, t1+50, t1+100)), 
               RasterIO = list(nXOff = c(1), 
                               nYOff = c(1), 
                               nXSize = 50, 
                               nYSize = 50, 
                               bands = c(1:6)))
```

@@

I can rename it like this
```{r}
x <- setNames(x, "Attribute_A")
```

@@

And time is four dates covering four months in 2018:

@@

### Calculate reference values
#### Import

@@

```{r}
names(x)
```
```{r}
x_sub <- x %>% 
  select(Attribute_A)
names(x_sub)
```

@@

Let's try it on the dummy data. Remember this data had three time steps.
```{r}
st_get_dimension_values(x, "time")
```

@@

```{r}
start <-  ymd("2018-06-01")
end <-  ymd("2018-08-31")
summer <- interval(start, end)
```

@@

Let me introduce some differences to Attribute_A between these two dates.
```{r}
x_aug$Attribute_A[,,,2] <- x_aug$Attribute_A[,,,1]*4
```

@@

For this example I want to calculate the mean temperature over the summer.
I therefore need to aggregate over time. Here is how:

@@

The time (and band) dimension is now gone as we have aggregated accoss it.

@@

Actually, this second method is better because we don't ned to filter out additional dimensions, and it retains the attribute name (the first method replaced it with the function name _mean_).

@@

>>>>

```{r}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
t1 = as.Date("1970-05-31")
x = read_stars(c(tif, tif, tif, tif), along = 
                  list(time = c(t1, t1+1, t1+50, t1+100)), 
               RasterIO = list(nXOff = c(1), 
                               nYOff = c(1), 
                               nXSize = 50, 
                               nYSize = 50, 
                               bands = c(1:6)))
```

@@

I can rename it like this

```{r}
x <- setNames(x, "Attribute_A")
```

@@

And time is four dates covering four months in 1970:

@@

### Calculate reference values

#### Import

@@

```{r}
x_sub <- x %>% 
  select(Attribute_A)
names(x_sub)
```

@@

Let's try it on the dummy data. Remember this data had three time steps.

@@

```{r}
start_month <-  "06"
end_month <- "08"
year_temp <- year(st_get_dimension_values(x, "time")[1])
start <- ym(paste(year_temp, start_month, sep="-"))
end <-  ym(paste(year_temp, end_month, sep="-"))
summer <- interval(start, end)
```

@@

Let me introduce some differences to Attribute_A between these two dates.

```{r}
x_aug$Attribute_A[,,,2] <- x_aug$Attribute_A[,,,1]*4
```

@@

For this example I want to calculate the mean temperature over the summer. I therefore need to aggregate over time. Here is how.

*Method 1:*

@@

The time (and band) dimension is now gone as we have aggregated across it.

*Method 2:*

@@

Actually, this second method is better because we don't need to filter out additional dimensions, and it retains the attribute name (the first method replaced it with the function name *mean*).

@@

#### Aggregate across years

For this step we need some more dummy data from different years.

```{r}
oldTime <- st_get_dimension_values(x_summerMean_v2, "time")

y1 <- x_summerMean_v2
y2 <- x_summerMean_v2 %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% mutate(Attribute_A = Attribute_A*rnorm(1,1,.3)) 
y3 <- x_summerMean_v2 %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% mutate(Attribute_A = Attribute_A*rnorm(1,1,.3))
y4 <- x_summerMean_v2 %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% mutate(Attribute_A = Attribute_A*rnorm(1,1,.3))
y5 <- x_summerMean_v2 %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% mutate(Attribute_A = Attribute_A*rnorm(1,1,.3))

# We combine the data into one cube:
y <- c(y1, y2, y3, y4, y5)
```

```{r}
ggplot() + 
  geom_stars(data = y) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~time)
```

```{r, eval=F, include=F}
# I would normally do the above code chink using tmap, but I cant get it to work when faceting across dimensions:
tm_shape(y)+
  tm_raster()+
  tm_facets("time")
```

Now we aggregate across years. For the reference period is is probably best to use *Method 1*, and for he indicator variable (the moving window estimation) probably *Method 2* is best. So Method 1 now.

```{r}
median_sd <- function(x) { c(median = median(x), sd = sd(x))}
```

```{r}
y_ref <- y %>%
  filter(band==1) %>%
  st_apply(c("x", "y"), FUN =  median_sd)
```

```{r}
dim(y_ref)
```

```{r}
st_get_dimension_values(y_ref, "median_sd")
```

It's easier if median and sd are unique attributes, rather than levels of a dimension.

```{r}
y_ref <- y_ref %>% split("median_sd")
```

The attribute name *mean* we can change this to something more meaningful.

```{r}
y_ref <- setNames(y_ref, c("reference_value", "sd"))
```

```{r}
tmap_arrange(
tm_shape(y_ref)+
  tm_raster("reference_value")
,
tm_shape(y_ref)+
  tm_raster("sd",
            palette = "-viridis")
)
```

#### Calculate variable values

The climate variable is the same as y1, y2, y~*n*~, ....
But to create some differences between the variable and the reference values, I will tweak these slightly, more or less divide them by two (pluss some error)

```{r}
y1v <- y1 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop() 
y2v <- y2 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop() 
y3v <- y3 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop() 
y4v <- y4 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop() 
y5v <- y5 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop()

#y_v <- c(y1v, y2v, y3v, y4v, y5v, y_ref)

y_v <- c(y1v, select(y_ref, reference_value))
y_v_merged <- merge(y_v)
dim(y_v_merged)
names(y_v_merged)
y_v_merged <- setNames(y_v_merged, "value")

st_get_dimension_values(y_v_merged, "attributes")
```

#### Normalise variable

```{r}
system.time(
  temp <- st_apply(y_v_merged, c("x", "y"), FUN = function(Attribute_A,reference_value) Attribute_A/reference_value)
)
```

```{r}
tm_shape(temp)+
  tm_raster("value")
```

```{r}

y_scaled <- merge(y_v)
dim(y_scaled)
names(y_scaled)
```

```{r}
tm_shape(y_v_scaled)+
  tm_raster()
```
